Typesafe Coordinate Systems in High-Throughput Sequencing Applications

### 2022-09-13 manuscript/main.tex
Tweak line breaks

While  sequence graphs will play an increasingly important role in the future,\cite{Paten01052017} the vast majority of sequence analysis work presently operates on coordinate systems with respect to a linear reference. These coordinate systems are defined on the one-dimensional axis of the reference sequence (which is given an integral or string identifier), with sequence intervals describing a range of positions within the reference. These systems inherently rely on two characteristics: the \textit{basis} (a zero or one indexed counting of the reference sequence), and---when considering intervals---by \textit{openness}: whether the end coordinate of a range is included or excluded (\textit{i.e.}, whether the interval is closed or open). The Cartesian product of the 2 values each of basis and openness yields four distinct potential coordinate systems, which are illustrated in Figure \ref{fig:coordinatesystems}.

@@

The reference implementation of Typesafe Coordinates as part of \textit{dhtslib}\cite{gregory_dhtslib_nodate} --- to our knowledge the first such polymorphic implementation of this concept in any bioinformatics tooling --- is written in D (\url{https://dlang.org/}), a statically-typed, compiled language. We found D's compile-time introspection and metaprogramming features ideal for construction of such a system. We also implemented Typesafe Coordinates in other languages, including Rust, OCaml, and Python, albeit with language-specific limitations in each of these.

@@

The authors wish to thank Brent Pedersen and Nicholas Wilson for critically reading the draft manuscript, and Seb Mondet for reviewing the OCaml code.

>>>>

While  sequence graphs will play an increasingly important role in the future,\cite{Paten01052017} the vast majority of sequence analysis work presently operates on coordinate systems with respect to a linear reference. These coordinate systems are defined on the one-dimensional axis of the reference sequence (which is given an integral or string identifier), with sequence intervals describing a range of positions within the reference. These systems inherently rely on two characteristics: the \textit{basis} (a zero or one indexed counting of the reference sequence), and---when considering intervals---by \textit{openness}: whether the end coordinate of a range is included or excluded (\textit{i.e.}, whether the interval is closed or open). The Cartesian product of the 2 values each of basis and openness yields four distinct potential coordinate systems (Figure \ref{fig:coordinatesystems}).

@@

The reference implementation of Typesafe Coordinates as part of \mbox{\textit{dhtslib}\cite{gregory_dhtslib_nodate}} --- to our knowledge the first such polymorphic implementation of this concept in any bioinformatics tooling --- is written in D (\url{https://dlang.org/}), a statically-typed, compiled language. We found D's compile-time introspection and metaprogramming features ideal for construction of such a system. We also implemented Typesafe Coordinates in other languages, including Rust, OCaml, and Python, albeit with language-specific limitations in each of these.

@@

The authors thank Brent Pedersen and Nicholas Wilson for critically reading the draft manuscript, and Seb Mondet for reviewing the OCaml code.
